Simulating High-Dimensional Multivariate Data

### 2021-03-01 020-simulation-algorithm.Rmd
Revise 020 prior to CSDA submission

The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceedings as follows:

@@

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic. In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$. Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,\ldots, d, \; i \neq j$. For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies. For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].

To maintain scalability (SP1), our solution is to essentially avoid this complication in Pearson matching. Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures, such as $\rho_S$ and $\tau$, from a modeling standpoint. Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), preserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above. This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed below in Section \@ref(rand-vec-gen)).

Now, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution [@Song00] in many practical applications. The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate. Later, we study robustness of our method to this limitation in selected [Monte Carlo evaluations]({#simulations}).

@@

@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. They found for what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

@@

1. Mapping step
    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.

    or

    (i$^\prime$) Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
2. Check admissibility and, if needed, compute the nearest correlation matrix.
    (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.
	Otherwise,
    (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.
3. NORTA step
    (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
    (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

The "mapping step" takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`).

Our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices. The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN.

In particular, if we let $X$ be MVN with $d$ components and set

@@

then there are 1-1 mappings between these sets. We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive. In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal. We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$) and also for checking that each pairwise correlation value is admissible (via the MC estimation of the 2-dimensional Frechet bounds).

Yet this scheme still suffers similar issues as the original Pearson-matching NORTA algorithm regarding the increasing probability of encountering a non-admissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not in $\Omega_P$ (or even a correlation matrix). The second step handles this issue by computing the nearest correlation matrix to $R_X$. In Section \@ref(package) we provide a basic description of how to quantify and potentially control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

The final step implements a high-performance NORTA algorithm. Specifically, step 3i uses a fast multi-core multivariate normal simulator, and steps 3ii, 3iii are distributed over CPU cores.

## Algorithm details

*Pearson Matching*
Given a desired target correlation $\rho_Y$ between a pair of marginal distributions, the correlation coefficient $\rho_Z$ used in the Gaussian copula must be specified so as to satisfy equation \@ref(eq:pearsonIntegralRelation).

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j
    (\#eq:pearsonIntegralRelation)
\end{equation}

where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables. In general, the double integral in \@ref(eq:pearsonIntegralRelation) is difficult or impossible to solve analytically, so numerical methods are required. Our implementation is based off the work of @xiao2019matching where the marginal transformation is approximated by a weighted sum of Hermite polynomials.

*Nearest Correlation Matrix*

This algorithm is trying to find a correlation matrix $X$ close to $G$ in the convex optimization problem:

* Qi, H., & Sun, D. (2006). A quadratically convergent Newton method for computing the nearest correlation matrix. SIAM journal on matrix analysis and applications, 28(2), 360-385.

>>>>

The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:

@@

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is computationally costly. In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$. Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
    (\#eq:pearsonIntegralRelation)
\end{equation}

where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables. For HD data, (nearly) exact evaluation may become too costly to enable practical simulation studies. For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely [@Xia17] and the publicly available `nortaRA` R package [@Chen2001]. As described later, to enable HD Pearson matching we approximate these integrals.

@@

@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

@@

1. Mapping step
    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
    (i') Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$. Alternatively,  
    (i'') Convert the target Pearson correlation matrix to $R_P$ to the corresponding, approximate MVN Pearson correlation $R_X$.  
2. Check admissibility and, if needed, compute the nearest correlation matrix.
    (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
	Otherwise,
    (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.  
3. Gaussian copula
    (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz. $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
    (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

*Step 1*. The first two descriptions of the *Mapping step* employ the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) and \@ref(eq:convertSpearman), respectively (implemented as `cor_covert`). Initializing our algorthim to match the nonparametric correlations by computing these equations for all pairs is computationally trivial. For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins. To this end, we implement the approximation scheme introduced by @xiao2019matching. Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approimated by weighted sums of Hermite polynomials.

*Step 2*. Once $R_X$ has been determined, we check admissibility of the target correlation via the steps described above. If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix. Finding the nearest correlation matrix is a common statistical computing problem. The defacto default function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing. As implemented the function fails to scale to HD.  Instead, we provide the quadratically convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically). The nearest correlation matrix problem can be written down as the following convex optimization problem:

For nonparametric correlation measures, our algorithm allows the generation of high-dimensional multivariate data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall $\tau$ matrices. The admissible classes consist of the matrices that map to a Pearson correlation matrix for a MVN. In particular, if we let $X$ be MVN with $d$ components and set

@@

then there are 1-1 mappings between these sets. We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive. In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal. We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$).

There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix. In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

*Step 3* The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins. Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation. Then all that is required is a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals. The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\phi(\cdot)$. The quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.
